---
title: "Creating Plugins"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Creating Plugins}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(rpp)
```

This documents how to create plugins for {rpp} so you can
extend its functionality.
Here we do so by building a small plugin to allow switching
between the {magrittr} pipe and R's native pipe.
It's a good use case, the native pipe does require additional
dependencies but only works with R >= 4.1.0.

## Creating the Plugin

The only rpp function we need is `inline_plugin()`.
This function only accepts two arguments: `dev` and `prod`,
both expect callback functions.
These callback functions should accept a single argument:
a vector of character strings--the source code where
each character string is a line of code.
For future extensibility they should also accept an ellipsis `...`.

First, let's create the function that will replace the native pipe
to magrittr and vice versa: the callback functions.

```{r}
chg_magrittr_to_native <- function(lines, ...) {
	gsub("\\%>\\%", "|>", lines)
}

chg_native_to_magrittr <- function(lines, ...) {
	gsub("\\|>", "%>%", lines)
}
```

The argument `lines` expects the vector of lines of source code,
`gsub()` is vectorised and simply replaces the pipes.
Note that this is very basic and will likely not work every where
correctly due to its simplicity (it's currently contextless).

We can nonetheless test this function.

```{r}
# sample code
magrittr <- 'cars %>% filter(dist > 60) %>% select(speed)'

# convert to native
native <- chg_magrittr_to_native(magrittr)
native
# convert back to magrittr
chg_native_to_magrittr(native)
```

## Exporting the plugin

This works, but to be useful we have to integrate it with rpp.
Within a package, create the following two functions to switch
between pipes, one of these functions does it one way and the other
the opposite way.

```r
#' @export 
rpp_magrittr_to_native <- function() {
  rpp::inline_plugin(
		dev = chg_magrittr_to_native, 
		prod = chg_magrittr_to_native
	)
}

#' @export 
rpp_native_to_magrittr <- function() {
  rpp::inline_plugin(
		dev = chg_native_to_magrittr, 
		prod = chg_native_to_magrittr
	)
}
```

## Using the plugin

To use this within a package add your function of choice
to the `DESCRIPTION`.

```
Config/rpp/plugins: list(myPkg::rpp_native_to_magrittr)
```

Then toggle between pipes with:

```r
rpp::rpp_to_dev()
rpp::rpp_to_prod()
```

These effectively run the respective dev and prod functions
on the source code in the `R/` directory.


## roxygen2 integration

The rpp package implements two roxygen2 roclets:

- `rpp::rpp_to_dev()` switches to dev mode
- `rpp::rpp_to_prod()` switches to prod mode

This serves two purposes:

1. The `rpp_to_prod()` roclet can be configured to be run with `devtools::document()`.
2. In the future, plugins will receive additional context information from roxygen2. This can be used to e.g. trigger additional behaviors or to adapt the documentation of function argument and return types.

Currently, plugins receive the contents of entire source files.
Later, plugins will be called with the contents of individual roxygen2 chunks.
We strive to keep the interfaces backward-compatible so that plugins developed today will remain operational for some time in the future.
